---
title: "Brain targeted metabolomic for CP2 project"
output: html_document
author: "Esraa Gabal - Baloni Lab - Purdue University"
date: "2025-11-18"
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_knit$set(root.dir = "/Users/eso1993/Library/CloudStorage/Box-Box/CP2_project/brain_targeted_metabolomics")
```

## load the required packages and read the data 
```{r, message=FALSE, error=FALSE, warning=FALSE}
# installing MetaboAnalystR
# install necessary pacakages and dependcies 
# install.packages("Rcpp")
# install.packages("RcppArmadillo")
# install.packages("qs")
# install.packages("Cairo")
# install.packages("devtools")
# install.packages("pacman")
library(pacman)
pacman::p_load(c("impute", "pcaMethods", "globaltest", "GlobalAncova", "Rgraphviz", "preprocessCore", "genefilter", "SSPA", "sva", "limma", "KEGGgraph", "siggenes","BiocParallel", "MSnbase", "multtest", "RBGL", "edgeR", "fgsea"))
# install.packages(c("BiocManager", "Rcpp", "devtools"))
# BiocManager::install(c("Biobase", "pcaMethods", "siggenes", "MSnbase", "impute", "fgsea", "edgeR", "BiocParallel", "RBGL"))
Sys.setenv(LD_LIBRARY_PATH=paste(Sys.getenv("LD_LIBRARY_PATH"), "/opt/X11/lib", sep=":"))
# devtools::install_local("/Users/eso1993/MetaboAnalystR/", upgrade = "never", dependencies = TRUE, build_vignettes = FALSE)

library(here)
library(dplyr)
library(purrr)
library(tidyr)
library(tidyverse)
library(ggplot2)
library(tidyplots)
library(scales)
library(MetaboAnalystR)
library(iheatmapr)
library(viridis)
library(RColorBrewer)
library(pheatmap)

# read the files 
tca <- read.csv(here("processed_files_CSV", "TCA.csv"))
acylcar <- read.csv(here("processed_files_CSV", "Acylcarnitines.csv"))
nefa <- read.csv(here("processed_files_CSV", "NEFA_Long_chain_FA.csv"))
scfa <- read.csv(here("processed_files_CSV", "SCFA_Short_chain_FA.csv"))
purv_CoA <- read.csv(here("processed_files_CSV", "Pyruvate_AcetylCoA.csv"))
```

## visualizing the measurments of all metabolites (TCA mets)
```{r error=FALSE, fig.align='center', fig.dpi=600, message=FALSE, fig.width=12, fig.height=6, warning=FALSE}
tca_long <- tca %>%
  mutate(Group = factor(Group,
                        levels = c("NTG-Veh", "APPPS1-Veh", "APPPS1-CP2"))) %>%
  pivot_longer(
    cols = 4:14,
    names_to  = "Metabolite",
    values_to = "Value")

# customed colors
my_colors <- new_color_scheme(c("#9BCD9B", "#104E8B", "#FFA54F"), name = "new_color")

# applied Dunn test method for performing Kruskal-Wallis test and adjusting p values using Benjamin Hocheberg method 

p <- tca_long |> 
  tidyplot(x = Group, y = Value, color = Group) |> 
  add_mean_bar(alpha = 0.7) |> 
  add_sem_errorbar(color = "black") |> 
  add_data_points_beeswarm(color = "black", size = 0.5) |> 
  adjust_colors(new_colors = my_colors) |>
  remove_x_axis_title() |>
  add_test_asterisks(method = "dunn_test", hide_info = TRUE, p.adjust.method = "BH") |> 
  adjust_x_axis(rotate_labels = 45, labels = c("NTG-Veh", "APP/PS1-Veh", "APP/PS1-CP2")) |>
  adjust_y_axis_title("Concentration (nmol/µg)") |>
  adjust_title(fontsize = 10, face = "bold") |>
  split_plot(by = Metabolite) 
p
```

## Acylcarnitines 
```{r, message=FALSE, error=FALSE, fig.dpi=600, fig.align='center', fig.width=12, fig.height=8, warning=FALSE}
acyl_long <- acylcar %>%
  mutate(Group = factor(Group,
                        levels = c("NTG-Veh", "APPPS1-Veh", "APPPS1-CP2"))) %>%
  pivot_longer(cols = 4:17,
    names_to  = "Metabolite",
    values_to = "Value")

# customed colors
my_colors <- new_color_scheme(c("#9BCD9B", "#104E8B", "#FFA54F"), name = "new_color")

p <- acyl_long |> 
  tidyplot(x = Group, y = Value, color = Group) |> 
  add_mean_bar(alpha = 0.7) |> 
  add_sem_errorbar(color = "black") |> 
  add_data_points_beeswarm(color = "black", size = 0.5) |> 
  adjust_colors(new_colors = my_colors) |>
  remove_x_axis_title() |>
  add_test_asterisks(method = "dunn_test", hide_info = TRUE, p.adjust.method = "BH") |> 
  adjust_x_axis(rotate_labels = 45, labels = c("NTG-Veh", "APP/PS1-Veh", "APP/PS1-CP2")) |>
  adjust_y_axis_title("Concentration (pmol/µg)") |>
  adjust_title(fontsize = 10, face = "bold") |>
  split_plot(by = Metabolite) 
p

```

## Long-chain fatty acids (NEFA)
```{r, message=FALSE, error=FALSE, fig.dpi=600, fig.align='center', fig.width=12, fig.height=6, warning=FALSE}
nefa_long <- nefa %>%
  mutate(Group = factor(Group,
                        levels = c("NTG-Veh", "APPPS1-Veh", "APPPS1-CP2"))) %>%
  pivot_longer(cols = 4:15,
    names_to  = "Metabolite",
    values_to = "Value")

# customed colors
my_colors <- new_color_scheme(c("#9BCD9B", "#104E8B", "#FFA54F"), name = "new_color")

p <- nefa_long |> 
  tidyplot(x = Group, y = Value, color = Group) |> 
  add_mean_bar(alpha = 0.7) |> 
  add_sem_errorbar(color = "black") |> 
  add_data_points_beeswarm(color = "black", size = 0.5) |> 
  adjust_colors(new_colors = my_colors) |>
  remove_x_axis_title() |>
  add_test_asterisks(method = "dunn_test", hide_info = TRUE, p.adjust.method = "BH") |> 
  adjust_x_axis(rotate_labels = 45, labels = c("NTG-Veh", "APP/PS1-Veh", "APP/PS1-CP2")) |>
  adjust_y_axis_title("Concentration (pmol/µg)") |>
  adjust_title(fontsize = 10, face = "bold") |>
  split_plot(by = Metabolite) 
p
```

## Short-chain fatty acids (SCFA)
```{r, message=FALSE, error=FALSE, fig.dpi=600, fig.align='center', fig.width=12, fig.height=6, warning=FALSE}
scfa_long <- scfa %>%
  mutate(Group = factor(Group,
                        levels = c("NTG-Veh", "APPPS1-Veh", "APPPS1-CP2"))) %>%
  pivot_longer(cols = 4:11,
    names_to  = "Metabolite",
    values_to = "Value")

# customed colors
my_colors <- new_color_scheme(c("#9BCD9B", "#104E8B", "#FFA54F"), name = "new_color")

p <- scfa_long |> 
  tidyplot(x = Group, y = Value, color = Group) |> 
  add_mean_bar(alpha = 0.7) |> 
  add_sem_errorbar(color = "black") |> 
  add_data_points_beeswarm(color = "black", size = 0.5) |> 
  adjust_colors(new_colors = my_colors) |>
  remove_x_axis_title() |>
  add_test_asterisks(method = "dunn_test", hide_info = TRUE, p.adjust.method = "BH") |> 
  adjust_x_axis(rotate_labels = 45, labels = c("NTG-Veh", "APP/PS1-Veh", "APP/PS1-CP2")) |>
  adjust_y_axis_title("Concentration (nmol/µg)") |>
  adjust_title(fontsize = 10, face = "bold") |>
  split_plot(by = Metabolite) 
p
```

## Pyruvate
```{r, message=FALSE, error=FALSE, fig.dpi=600, fig.align='center', warning=FALSE}
pyruvate <- purv_CoA[,-4] # get only the info for pyruvate
pyruvate$Group <- factor(pyruvate$Group, levels = unique(pyruvate$Group)) # get the groups as factors of three levels

p <- pyruvate |> 
  tidyplot(x = Group, y = `Pyruvate..nmol.µg.`, color = Group) |> 
  add_mean_bar(alpha = 0.7) |> 
  add_sem_errorbar(color = "black") |> 
  add_data_points_beeswarm(color = "black", size = 0.5) |> 
  adjust_colors(new_colors = my_colors) |>
  remove_x_axis_title() |>
  add_test_asterisks(method = "dunn_test", hide_info = TRUE, p.adjust.method = "BH") |> 
  adjust_x_axis(rotate_labels = 45, labels = c("NTG-Veh", "APP/PS1-Veh", "APP/PS1-CP2")) |>
  adjust_y_axis_title("Concentration (nmol/µg)") |>
  adjust_title(fontsize = 10, face = "bold")
p
```

## Acetyl-CoA
```{r, message=FALSE, error=FALSE, fig.dpi=600, fig.align='center', warning=FALSE}
coA <- purv_CoA[,-3] # get only the info for acetyl-CoA
coA$Group <- factor(coA$Group, levels = unique(coA$Group)) # get the groups as factors of three levels

p <- coA |> 
  tidyplot(x = Group, y = `acetyl.CoA..pmol.µg.`, color = Group) |> 
  add_mean_bar(alpha = 0.7) |> 
  add_sem_errorbar(color = "black") |> 
  add_data_points_beeswarm(color = "black", size = 0.5) |> 
  adjust_colors(new_colors = my_colors) |>
  remove_x_axis_title() |>
  add_test_asterisks(method = "dunn_test", hide_info = TRUE, p.adjust.method = "BH") |> 
  adjust_x_axis(rotate_labels = 45, labels = c("NTG-Veh", "APP/PS1-Veh", "APP/PS1-CP2")) |>
  adjust_y_axis_title("Concentration (pmol/µg)") |>
  adjust_title(fontsize = 10, face = "bold")
p
```

## prepare all metabolites for MetaboanalystR
```{r, message=FALSE, warning=FALSE, error=FALSE, fig.height=6, fig.width=12}

# exclude the metabolites with zero concentration
# convert nmol to pmol so that all measumrents are unified for certain mets with nmol conc
tca_new <- tca %>%
  select(-Unit) %>%
  mutate(across(where(is.numeric), ~ . * 1000))

scfa_new <- scfa %>%
  select(-c(Unit,Group)) %>%
  mutate(across(where(is.numeric), ~ . * 1000))

pyrv_coa_new <- purv_CoA %>%
  select(-Group) %>%
  mutate(across(2, ~ . * 1000))
colnames(pyrv_coa_new) <- c("SampleID", "Pyruvate", "Acetyl-CoA")

# combine all metabolites in one dataframe 
met_all <- list(tca_new, acylcar[,-c(2,3,11)], nefa[,-c(2,3,10,14)], scfa_new, pyrv_coa_new) %>%
  reduce(full_join, by = "SampleID")

# modify the column names 
colnames(met_all) <- gsub("_", " ", colnames(met_all))
colnames(met_all) <- gsub("\\.", "-", colnames(met_all))
colnames(met_all) <- sapply(colnames(met_all), function(x) {paste0(toupper(substring(x, 1, 1)), substring(x, 2))})

#write.csv(met_all, "/Users/eso1993/Library/CloudStorage/Box-Box/CP2_project/brain_targeted_metabolomics/processed_files_CSV/all_metabolites.csv", row.names = FALSE)
```

## Metaboanalyst R
## Proceed data for MetaboanalystR and perform sanity check
```{r, message=FALSE, error=FALSE, warning=FALSE}
rm(list = ls())
# initialize the work with MetaboAnalystR
# create the object of MetaboAnalystR
mSet<-InitDataObjects("conc", "stat", FALSE)

# read the data file by determining its path 
mSet<-Read.TextData(mSet, here("processed_files_CSV/all_metabolites.csv"))
mSet<-SanityCheckData(mSet) # check the data 
```

## Univariate Analysis
```{r, message=FALSE, warning=FALSE, error=FALSE, fig.height=6, fig.width=12}
#Mean centering scaling as it is the most traditional method for pretreating metabolomics data. Centring is used to correct for the differences between high and low abundant metabolites, allowing the data analysis to focus on the mean of the metabolite concentrations,

mSet<-ReplaceMin(mSet)
mSet<-PreparePrenormData(mSet)

mSet<-Normalization(mSet,"NULL", "LogNorm", "MeanCenter", ratio=FALSE, ratioNum=FALSE)

# row-wise normalization, followed by trsansformation and scaling of the data 
# plotting normalized values of each metabolite across the samples 
mSet<-PlotNormSummary(mSet, "~/Desktop/brain-normalization_metabolomics", format ="png", dpi=600, width=NA)

# column-wise normalization, followed by transformation and scaling of the data 
# plotting the normalized values of all samples 
mSet<-PlotSampleNormSummary(mSet, "~/Desktop/brain-Normalizaton_accross_samples_metabolomics", format = "png", dpi=600, width=NA)
```

## Volcano analysis 
```{r, message=FALSE, warning=FALSE, error=FALSE, fig.width=20, fig.height=10, fig.align='center'}
# we will set the lfc threshold at 1.0 and a p-value cutt of 0.05, unequal variance (FALSE), and raw to use raw p-value
mSet<-Volcano.Anal(mSet, paired = FALSE, fcthresh = 1.0, 0, F, 0.05, FALSE, "raw")

```

## Discriminative anaylsis (PLS-DA)
```{r, message=FALSE, warning=FALSE, error=FALSE}
mSet <- PLSR.Anal(mSet, reg=TRUE)
mSet <- PlotPLSPairSummary(mSet, "~/Desktop/brain-PLS_PAIR_summary", "png", 600, width=NA, 5) # for 5 components to compare
mSet <- PlotPLS2DScore(mSet, "~/Desktop/brain-PLS_DA_2D_discriminative_analysis", "png", 600, width=NA, 1,2,0.95,1,0) # for only component 1 and 2 
# plot the top 10 important metabolites accross the samples 
mSet<-PlotPLS.Imp(mSet, "~/Desktop/brain-Important_metabolites_PLS_DA_Component1", "png", 600, width=NA, "vip", "Comp. 1", 10, FALSE) # for component 1
mSet<-PlotPLS.Imp(mSet, "~/Desktop/brain-Important_metabolites_PLS_DA_Component2", "png", 600, width=NA, "vip", "Comp. 2", 10, FALSE) # for component 2
```

## Sparse Partial Least Squares - Discriminant Analysis (sPLS-DA)
```{r, message=FALSE, error=FALSE, warning=FALSE}
mSet<-SPLSR.Anal(mSet, 5, 10, "same", "Mfold", 5, T) # plotting for the top 10 metabolites  

# Plot sPLS-DA overview
mSet<-PlotSPLSPairSummary(mSet, "~/Desktop/brain_sPLS-DA_pair_", format = "png", dpi=600, width=NA, 5) # 5 components

# Create 2D sPLS-DA Score Plot
mSet<-PlotSPLS2DScore(mSet, "~/Desktop/brain_sPLS-DA_2D_plot_", format = "png", dpi=600, width=NA, 1, 2, 0.95, 1, 0) # 2 components

# Create sPLS-DA loadings plot
mSet<-PlotSPLSLoading(mSet, "~/Desktop/brain_sPLS-DA_loading_top10_mets", format = "png", dpi=600, width=NA, 1,"overview")

# check for error 
mSet<-PlotSPLSDA.Classification(mSet, "~/Desktop/brain_sPLS_error_", format = "png", dpi=600, width=NA)
# error is high at all components plotting so we will go for the 3D plot with the least error detected

# ploting 3D plot for PC1,2,3 at angle 72
mSet<-PlotSPLS3DScoreImg(mSet, "~/Desktop/brain_sPLS-DA_3D_plot_", format = "png", 600, width=NA, 1, 2, 3, 72)

```

## Orthogonal Partial Least Squares - Discriminant Analysis (oPLS-DA)
```{r, message=FALSE, error=FALSE, warning=FALSE}
# Perform oPLS-DA analysis
mSet<-OPLSR.Anal(mSet, reg=TRUE)
# Create a 2D oPLS-DA score plot
mSet <- PlotOPLS2DScore(mSet, "~/Desktop/brain_oPLS_DA_2D", format = "png", dpi=600, width=NA, 1,2,0.95,1,0)
# Create a significant features plot
mSet <- PlotOPLS.Splot(mSet, "~/Desktop/brain_oPLS_DA_significant_features_plot", "all", "png", 600, width=NA)
# Create a plot of 20 features ranked by importance
mSet <- PlotOPLS.Imp(mSet, "~/Desktop/brain_oPLS_DA_ranked_feature_importance", "png", 600, width=NA, "vip", "tscore", 20,FALSE)
# Perform and plot oPLS-DA permutation 
mSet<-OPLSDA.Permut(mSet, 10000)
```

```{r}
# plot the oPLS-DA imputation test 
mSet<-PlotOPLS.Permutation(mSet, "~/Desktop/brain-oPLSDA-imputation_test_", format = "png", dpi=600, width=NA)
```

## Hierarchical Clustering: Heatmaps
```{r, warning=FALSE, error=FALSE, message=FALSE, fig.height=13, fig.width=18}
# we will proceed heatmaps using the z-score normalized metabolites 
meta_all <- read.csv(here("processed_files_CSV/all_metabolites.csv"))

# reformat the column names 
colnames(meta_all) <- gsub("X", "", colnames(meta_all))
colnames(meta_all) <- gsub("\\.", " ", colnames(meta_all))

# z-normalise
meta_all_normalized <- scale(meta_all[-c(1,2)])
meta_all_normalized_new <- cbind(meta_all[,c(1:2)], meta_all_normalized)

# prepare the annotations
Group_annotation <- data.frame(SampleID = meta_all_normalized_new$SampleID, Group = meta_all_normalized_new$Group)
Group_annotation$Group <- factor(Group_annotation$Group, levels = unique(Group_annotation$Group))
rownames(Group_annotation) <- Group_annotation$SampleID

# prepare the data 
meta_all_normalized_new <- meta_all_normalized_new[,-2]
meta_all_normalized_new <- t(meta_all_normalized_new) # transpose to have the metabolites as rows
colnames(meta_all_normalized_new) <- meta_all_normalized_new[1,]
meta_all_normalized_new <- meta_all_normalized_new[-1,]

meta_all_numeric <- apply(meta_all_normalized_new, 2, as.numeric)
rownames(meta_all_numeric) <- rownames(meta_all_normalized_new)
colnames(meta_all_numeric) <- colnames(meta_all_normalized_new)
meta_all_numeric <- as.matrix(meta_all_numeric)

# ensure the order of samples is same
all(rownames(Group_annotation) == colnames(meta_all_numeric))

# plot heatmap
brewer_palette <- colorRampPalette(brewer.pal(5, "PRGn"))(100) 
plot <- pheatmap(meta_all_numeric, cluster_rows = T, show_rownames=T,
                 border_color="grey", scale="none",
                 height=45, annotation_col = Group_annotation, cluster_cols = F, angle_col ="45",
                 fontsize_col = 10, fontsize_row = 11,
                 cellwidth = 20, cellheight = 10, gaps_col = c(5,10), color = brewer_palette)

plot
#ggsave("~/Desktop/brain_metabolites-cp2.png", plot = plot, dpi = 600, bg = "white", width = 30, height = 25)
```
