---
title: "iMiceBrain analyses and visualisation"
author: "Esraa Gabal - Baloni Lab - Purdue University"
date: "2025-08-21"
output: html_document
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_knit$set(root.dir = "/Users/eso1993/Library/CloudStorage/Box-Box/")
```

## installing and activating required libraries
```{r, message=FALSE, warning=FALSE}
library(here)
library(dplyr)
library(tidyr)
library(tidyverse)
library(ggplot2)
library(tidyplots)
library(pheatmap)
library(VennDiagram)
library(grid)
library(futile.logger)
library(ggVennDiagram)
library(UpSetR)
library(gridExtra)
library(igraph)
library(ggraph)
library(biomaRt)
library(eulerr)
```

## Applying TPM and Limma normalization on brain public datasets before model integration

```{r}
TPM_brain <- read.csv(here("iMiceBrain_required_data", "TPM_normalized_brain.csv"))
limma_brain <- read.csv(here("iMiceBrain_required_data", "limma_normalized_brain.csv"))
raw_brain <- read.csv(here("iMiceBrain_required_data", "raw_brain_data.csv"))
brain_documentation <- read.csv(here("iMiceBrain_required_data", "brain_data_documentation.csv"))

# assign the brain regions as factor 
brain_documentation$Brain_tissue <- factor(brain_documentation$Brain_tissue, levels = unique(brain_documentation$Brain_tissue))

# plotting the TPM and Limma normalized samples considering brain regions for coloring
# convert to a long format
TPM_brain_longer <- TPM_brain %>%
  pivot_longer(cols = -Gene, names_to = "SampleID", values_to = "GeneCount")

limma_brain_longer <- limma_brain %>%
  pivot_longer(cols = -Gene, names_to = "SampleID", values_to = "GeneCount")

raw_brain_longer <- raw_brain %>%
  pivot_longer(cols = -Gene, names_to = "SampleID", values_to = "GeneCount")


# inspect structure to ensure numeric values 
str(TPM_brain_longer)
str(limma_brain_longer)
str(raw_brain_longer)

# compute the log10 of gene count for TPM normalized and raw data 
TPM_brain_longer$log10Gene <- log10(TPM_brain_longer$GeneCount)
raw_brain_longer$log10Gene <- log10(raw_brain_longer$GeneCount)

# replcae NAs by zero
TPM_brain_longer[is.na(TPM_brain_longer)] <- 0
raw_brain_longer[is.na(raw_brain_longer)] <- 0

head(TPM_brain_longer, n = 5)
head(limma_brain_longer, n = 5)
head(raw_brain_longer, n = 5)

```

# boxplotting with colored regions
```{r, fig.width=20, fig.height=10}
# subset the brain documentation and asisgn a column names for sample ids matching that of count data 
brain_regions <- brain_documentation[,c("GSM_accession","Brain_tissue")]
colnames(brain_regions)[names(brain_regions) == 'GSM_accession'] <- "SampleID"

# merge to the count data 
TPM_brain_longer <- left_join(TPM_brain_longer, brain_regions, by = "SampleID") 
limma_brain_longer <- left_join(limma_brain_longer, brain_regions, by = "SampleID") 
raw_brain_longer <- left_join(raw_brain_longer, brain_regions, by = "SampleID") 

head(TPM_brain_longer, n = 5)
head(limma_brain_longer, n = 5)
head(raw_brain_longer, n = 5)

# plot 
ggplot(raw_brain_longer, aes(x = SampleID, y = log10Gene, fill = Brain_tissue)) +
  geom_boxplot() + theme_minimal() +
  labs(x = "Sample ID", y = "log10 Gene Count", title = "Raw (no-normalization) brain transcriptomics") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 9)) 

ggplot(TPM_brain_longer, aes(x = SampleID, y = log10Gene, fill = Brain_tissue)) +
  geom_boxplot() + theme_minimal() +
  labs(x = "Sample ID", y = "log10 Gene Count", title = "TPM normalized brain transcriptomics") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 9)) 

ggplot(limma_brain_longer, aes(x = SampleID, y = GeneCount, fill = Brain_tissue)) +
  geom_boxplot() + theme_minimal() +
  labs(x = "Sample ID", y = "Gene Count", title = "Limma normalized brain transcriptomics") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 9)) 

```

## comparison between brain-drafts and the consensus iMiceBrain
```{r, fig.width=25, fig.height=10}
genes <- read.csv(here("iMiceBrain_required_data", "mouseBrain_drafts_genes.csv"))
mets <- read.csv(here("iMiceBrain_required_data", "mouseBrain_drafts_metabolites.csv"))
rxns <- read.csv(here("iMiceBrain_required_data", "mouseBrain_drafts_reactions.csv"))

genes[] <- lapply(genes, as.character)

genes_long <- pivot_longer(genes, cols = everything(), names_to = "Model", values_to = "Count") %>% 
  mutate(Feature = "Genes")
mets_long <- pivot_longer(mets, cols = everything(), names_to = "Model", values_to = "Count") %>% 
  mutate(Feature = "Metabolites")
rxns_long <- pivot_longer(rxns, cols = everything(), names_to = "Model", values_to = "Count") %>% 
  mutate(Feature = "Reactions")

# combine 
combined_data <- bind_rows(genes_long, mets_long, rxns_long)

# remove any empty cells
combined_data[] <- lapply(combined_data, function(x) if (is.character(x)) trimws(x) else x)
combined_data <- combined_data %>%
  filter(!is.na(Count) & Count != "") %>%
  group_by(Model, Feature) %>%
  summarise(Count = n(), .groups = "drop")

# extract the model names only
combined_data$Model <- gsub("_.*", "", combined_data$Model)
combined_data$Model <- factor(combined_data$Model, levels = c(
  "mCADRE", "rFASTACOROMICS", "CORDA", "ftINIT"
))

# barplot visualisation
ggplot(combined_data, aes(x = Model, y = Count, fill = Feature)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8), color = "black") +
  labs(
    title = "Brain-Draft GEMs features",
    x = "Model",
    y = "Count"
  ) +
  scale_fill_brewer(palette = "PuBuGn") +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 15),
        axis.text.y = element_text(size = 15, margin = margin(r = 0.5)),
        axis.title.x = element_text(size = 16, face = "bold", margin = margin(t = 10)),
        axis.title.y = element_text(size = 16, face = "bold")) + ylim(c(0,10000))
```

## comparing the iMiceBrain with other models 
```{r}
iMiceBrain <- read.csv(here("iMiceBrain_required_data", "iMiceBrain_stats.csv"))
imm1865 <- read.csv(here("iMiceBrain_required_data", "iMM1865_stats.csv"))
recon3d <- read.csv(here("iMiceBrain_required_data", "Recon3d_stats.csv"))
brain_human <- read.csv(here("iMiceBrain_required_data", "Brain_atlas_harvey_model_stats.csv"))
```

```{r, fig.width=5, fig.height=5, warning=FALSE, message=FALSE}
# metabolites
met_list <- list(
  iMiceBrain = trimws(iMiceBrain$mets),
  iMM1865 = trimws(imm1865$mets),
  Recon3D = trimws(recon3d$mets),
  `Human brain` = trimws(brain_human$mets)
)

met_list <- lapply(met_list, function(x) x[!is.na(x) & x != ""]) # remove any NA

venn_met <- venn.diagram(
  x = met_list,
  category.names = c("iMiceBrain", "iMM1865", "Recon3D", "Human Brain"),
  filename = NULL,
  
  # Venn Diagram settings
  imagetype = "png",
  height = 3000,
  width = 3000,
  resolution = 600,
  compression = "lzw",
  
  # Circles
  lwd = 2,
  lty = 'blank',
  fill = c("#4B6C8B", "#8DA9C4", "#CBD2D0", "#E8E9EB"),
  
  # Numbers
  cex = 1.2,
  fontface = "bold",
  fontfamily = "sans",
  
  # Labels
  cat.cex = 0.9,
  cat.fontface = "bold",
  cat.default.pos = "outer",
  cat.pos = c(-30, 30, -30, 30),
  cat.dist = c(0.06, 0.06, 0.06, 0.06),
  cat.fontfamily = "sans",
  cat.col = "darkred")

grid.newpage()
grid.draw(venn_met)
```

```{r, fig.width=5, fig.height=5}
# reactions
rxns_list <- list(
  iMiceBrain = trimws(iMiceBrain$rxns),
  iMM1865 = trimws(imm1865$rxns),
  Recon3D = trimws(recon3d$rxns),
  `Human brain` = trimws(brain_human$rxns)
)

rxns_list <- lapply(rxns_list, function(x) x[!is.na(x) & x != ""]) # remove any NA

venn_rxn <- venn.diagram(
  x = rxns_list,
  category.names = c("iMiceBrain", "iMM1865", "Recon3D", "Human Brain"),
  filename = NULL,
  
  # Venn Diagram settings
  imagetype = "png",
  height = 3000,
  width = 3000,
  resolution = 600,
  compression = "lzw",
  
  # Circles
  lwd = 2,
  lty = 'blank',
  fill = c("#A67B5B", "#C9B79C", "#7C9473", "#DAD0C2"),
  
  # Numbers
  cex = 1.2,
  fontface = "bold",
  fontfamily = "sans",
  
  # Labels
  cat.cex = 0.9,
  cat.fontface = "bold",
  cat.default.pos = "outer",
  cat.pos = c(-30, 30, -30, 30),
  cat.dist = c(0.06, 0.06, 0.06, 0.06),
  cat.fontfamily = "sans",
  cat.col = "darkred")

grid.newpage()
grid.draw(venn_rxn)
```

## Checking the specificty of iMiceBrain genes using HPA and Allen mouse-brain data
```{r, fig.width=15, fig.height=10, warning=FALSE, message=FALSE}

##### HPA data 
mouse_brain <- read.csv(here("iMiceBrain_required_data", "transcript_rna_mousebrain_human_proteome_atlas.csv"))
mouse_brain <- mouse_brain[rowSums(mouse_brain[ , sapply(mouse_brain, is.numeric)]) >= 1.0, ] # include genes with total TPM is >1 a common threshold for conidering gene expressed in TPM

# get the entrez gene id using biomart
mart = useMart("ensembl", host = "https://www.ensembl.org", dataset = "mmusculus_gene_ensembl")
raw_id <- getBM(attributes = c("external_gene_name", "entrezgene_id"),
                filters = "external_gene_name",
                values = mouse_brain$Gene,
                mart = mart)

brain_id <- read.csv(here("iMiceBrain_required_data", "iMiceBrain_stats.csv"))
brain_id <- tibble(brain_id$genes)
brain_id <- na.omit(brain_id)
colnames(brain_id)[names(brain_id) == 'brain_id$genes'] <- "entrezgene_id"

raw_brain <- left_join(brain_id, raw_id, by= "entrezgene_id")
colnames(raw_brain)[names(raw_brain) == 'external_gene_name'] <- "Gene"

raw_brain_new <- left_join(raw_brain, mouse_brain, by = "Gene")
raw_brain_new <- na.omit(raw_brain_new)

# make a pie chart to show how many brain-specific genes covered in iMiceBrain vs the total metabolic genes captured in Human Proteome Atlas (HPA) mouse-brain transcriptomics

genes_covered <- intersect(raw_brain_new$entrezgene_id, raw_brain$entrezgene_id)

gene_coverage_counts <- c(
  `HPA Brain-Specific Genes` = length(genes_covered),
  `iMiceBrain-Specific Genes` = length(setdiff(raw_brain$entrezgene_id, genes_covered))
)

# Create data frame
df <- data.frame(
  group = names(gene_coverage_counts),
  count = gene_coverage_counts
)

# Add proportions and label positions
df <- data.frame(
  group = names(gene_coverage_counts),
  count = gene_coverage_counts
) %>%
  mutate(
    prop = count / sum(count) * 100,
    label = paste0(group, "\n", round(prop, 1), "%")
  )

# Compute manual ypos for better centering
df$ypos <- c(
  sum(df$prop) - df$prop[1] / 2,  
  df$prop[2] / 3                 
)

# Plot pie chart
pie1 <- ggplot(df, aes(x = "", y = prop, fill = group)) +
  geom_bar(stat = "identity", width = 2, color = "grey") +
  coord_polar("y", start = 0) +
  geom_text(aes(y = ypos, label = label), color = "black", size = 5) +
  scale_fill_brewer(palette = "Set4") +
  theme_void() +
  labs(title = "Brain-Specific Metabolic Genes in iMiceBrain", fill = "iMiceBrain genes") +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"))

##### allen data 
# read allen mouse-brain expression
allen <- read.table(here("iMiceBrain_required_data", "rna_mouse_brain_allen.tsv"), header = T, sep = "\t", quote = "")
allen$Gene <- tools::toTitleCase(tolower(allen$Gene))
allen$BrainRegion <- tools::toTitleCase(tolower(allen$BrainRegion))

# read the iMiceBrain model gene names 
ortho_imicebrain <- read_csv(here("iMiceBrain_required_data", "iMiceBrain_genes_human_orthologous.csv"))

# get the pie chart for percentage 
genes_covered_allen <- intersect(allen$Gene, ortho_imicebrain$mouseGene)

gene_coverage_counts_2 <- c(
  `Allen Brain-Specific Genes` = length(genes_covered_allen),
  `iMiceBrain-Specific Genes` = length(setdiff(ortho_imicebrain$mouseGene, genes_covered_allen))
)

# Create data frame
df_2 <- data.frame(
  group = names(gene_coverage_counts_2),
  count = gene_coverage_counts_2
)

# Add proportions and label positions
df_2 <- data.frame(
  group = names(gene_coverage_counts_2),
  count = gene_coverage_counts_2
) %>%
  mutate(
    prop = count / sum(count) * 100,
    label = paste0(group, "\n", round(prop, 1), "%")
  )

# Compute manual ypos for better centering
df_2$ypos <- c(
  sum(df_2$prop) - df_2$prop[1] / 2,  
  df_2$prop[2] / 3                 
)

# Plot pie chart
pie2 <- ggplot(df_2, aes(x = "", y = prop, fill = group)) +
  geom_bar(stat = "identity", width = 2, color = "grey") +
  coord_polar("y", start = 0) +
  geom_text(aes(y = ypos, label = label), color = "black", size = 5) +
  scale_fill_brewer(palette = "Set2") +
  theme_void() +
  labs(title = "Brain-Specific Metabolic Genes in iMiceBrain", fill = "iMiceBrain genes") +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"))

grid.arrange(pie1, pie2, ncol = 2)
```
# getting the iMiceBrain genes' expression from mouse-brain allen rna data 

```{r, fig.height=55, fig.width=15, warning=FALSE, message=FALSE}
# read the iMiceBrain model gene names 
ortho_imicebrain <- read_csv(here("iMiceBrain_required_data", "iMiceBrain_genes_human_orthologous.csv"))

# get the expression of iMiceBrain genes from the brain-mouse allen dataset
# extract the allen expression of those concerning iMiceBrain genes 
allen_vs_model <- allen %>% filter(Gene %in% ortho_imicebrain$mouseGene)
allen_vs_model <- allen_vs_model[,-1]

# convert to a wide format 
allen_vs_model_wide <- allen_vs_model %>%
  as.data.frame() %>%
  pivot_wider(
    names_from = BrainRegion,
    values_from = ExpressionEnergy,
    values_fill = 0  # fill missing with 0
  )
allen_vs_model_wide <- as.data.frame(allen_vs_model_wide)
rownames(allen_vs_model_wide) <- allen_vs_model_wide[,"Gene"]
allen_vs_model_wide <- as.matrix(allen_vs_model_wide[,-1])

# heatmap
pheatmap(allen_vs_model_wide, cluster_rows = F, show_rownames=F,
                      scale="row",
                      height=20, cluster_cols = T,
                      cellwidth = 50, cellheight = 2,
                      number_color = "black", color = hcl.colors(50, "BluYl"), angle_col = 45,
                      fontsize_col = 15, border_color = NA)
```


```{r, fig.width=5, fig.height=5, warning=FALSE}

# getting the barplots for gene number in each model
genes_all <- list(
  iMiceBrain   = trimws(iMiceBrain$genes),
  iMM1865      = trimws(imm1865$genes),
  Recon3D      = trimws(recon3d$genes),
  `Human brain` = trimws(brain_human$genes)
)

genes_all <- lapply(genes_all, function(x) x[!is.na(x) & x != ""])

# get the gene number 
gene_counts <- data.frame(
  Model = names(genes_all),
  GeneCount = sapply(genes_all, function(x) length(unique(x)))
)

ggplot(gene_counts, aes(x = Model, y = GeneCount)) +
  geom_bar(stat = "identity", fill = "grey70", color = "black", linetype = "dashed",
           size = 0.7, width = 0.5) +
  theme_minimal() +
  labs(title = "Gene Count per Model", x = "Model", y = "Unique Gene Count") +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.text.x = element_text(size = 12, face = "bold", margin = margin(r = 0.5)),
    axis.text.y = element_text(size = 12, margin = margin(t = 0.5)),
    text = element_text(size = 12)
  )
```

## Plotting the human-mouse orthologous genes in iMiceBrain vs human recon3d and brain-harvey
```{r, fig.width=20, fig.height=10, message=FALSE}

ortho_imicebrain <- read_csv(here("iMiceBrain_required_data", "iMiceBrain_genes_human_orthologous.csv"))

# getting the barplots for gene number in each model
genes_VENN <- list(
  iMiceBrain   = trimws(na.omit(ortho_imicebrain$humanID)),
  Recon3D      = trimws(na.omit(recon3d$genes)),
  `Human brain` = trimws(na.omit(brain_human$genes))
)

# get the gene number 
gene_counts <- data.frame(
  Model = names(genes_VENN),
  GeneCount = sapply(genes_VENN, function(x) length(unique(x)))
)

# please note that different gene ids in iMiceBrain for isozymes have the same human gene NCBI ID as captured by homologene and 
# hence they are have duplicates. the venn diagram captures only unique genes (27 duplicates in iMiceBrain)

venn_gene <- venn.diagram(
  x = genes_VENN,
  category.names = c("iMiceBrain", "Recon3D", "Human Brain"),
  filename = NULL,
  
  # Venn Diagram settings
  imagetype = "png",
  height = 3000,
  width = 3000,
  resolution = 600,
  compression = "lzw",
  
  # Circles
  lwd = 2,
  lty = 'blank',
  fill = c("khaki3", "indianred", "#8172B2"),
  
  # Numbers
  cex = 1.24,
  fontface = "bold",
  fontfamily = "sans",
  
  # Labels
  cat.cex = 1.4,
  cat.fontface = "bold",
  cat.default.pos = "outer",
  cat.pos = c(-30, 30, -30),
  cat.dist = c(0.02, 0.02, 0.02),
  cat.fontfamily = "sans",
  cat.col = "black")


# another way of mapping proportional Venn diagram

fit <- euler(
  list(
    iMiceBrain   = unique(genes_VENN$iMiceBrain),
    Recon3D      = unique(genes_VENN$Recon3D),
    Human_Brain  = unique(genes_VENN$`Human brain`)
  )
)

venn_2 <- plot(
  fit,
  fills = list(fill = c("khaki3", "indianred", "#8172B2"), alpha = 0.5),
  edges = TRUE,
  labels = list(
    labels = c("iMiceBrain", "Recon3D", "Human Brain"),
    font = 2
  ),
  quantities = list(cex = 1.3),  # numbers inside regions
  main = list(label = "Shared and Unique Genes", cex = 1.5, font = 2),
  legend = FALSE,
  margins = c(0.7, 0.7, 0.7, 0.7), cex = 1.4
)

grid.newpage()
venn_gene_grob <- grid.grabExpr(grid.draw(venn_gene))
venn_2_grob    <- grid.grabExpr(grid.draw(venn_2))

grid.arrange(venn_gene_grob, venn_2_grob, ncol = 2)
```

## getting the most abundant subsystems in iMiceBrain (top 20)
```{r, message=FALSE, warning=FALSE, fig.width=6.5, fig.height=6.6}
iMiceBrain = read.csv(here("iMiceBrain_required_data", "iMiceBrain_stats.csv"))

# summaries the number of subsystems and then plot the top 20 in a vertical barplot
iMiceBrain_sub <- iMiceBrain[, "subsystems", drop = FALSE]

iMiceBrain_summary <- iMiceBrain_sub %>%
  group_by(subsystems) %>%
  summarise(Total_Subsystems = n()) %>%
  arrange(desc(Total_Subsystems)) %>%
  slice_head(n = 20)

ggplot(iMiceBrain_summary, aes(x = reorder(subsystems, Total_Subsystems), y = Total_Subsystems)) +
  geom_bar(stat = "identity", fill = "#8B636C", color = "black") +
  coord_flip() +
  labs(title = "Top 20 Subsystems in iMiceBrain",
       x = "Subsystem",
       y = "Number of Reactions") +
  theme_minimal() + theme(axis.text.x = element_text(size = 13, margin = margin(t = 5)),
                          axis.text.y = element_text(size = 13, margin = margin(r = 5)),
                          axis.title.y = element_blank(),
                          axis.title.x = element_text(margin = margin(t = 10), size = 15))
```


